---
title: "Alberta Income Support: Historic Caseload Trends"
subtitle: "A literate R + Quarto pair — BFE/ETW breakdown analysis"
author: "Literate analysis — GitHub Copilot (Claude Sonnet 4.6)"
date: "April 2026"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 10
    fig-height: 5
    df-print: paged
    highlight-style: github
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Source the companion R script.
# All data loading, cleaning, and chart construction happen there;
# this document only adds narrative and calls the resulting objects.
source("analysis.R")
```

## Research Questions

This document is a reproducible, literate analysis of the Alberta Income Support
caseload dataset published by Alberta Assisted Living and Social Services (ALSS).
It directly addresses two questions:

1. **What is the historic trend of Income Support caseload?**
2. **How does breaking it down by BFE and ETW change the story?**

The companion script `analysis.R` contains every data-transformation step.
This file (`analysis.qmd`) provides the narrative, tables, and charts.

---

## Data Source

| Field | Value |
|:------|:------|
| **Publisher** | Alberta Assisted Living and Social Services (ALSS) |
| **Dataset** | Income Support Caseload |
| **Coverage** | April 2005 – September 2025 (246 calendar months) |
| **Licence** | [Open Government Licence – Alberta](https://open.alberta.ca/licence) |
| **Portal** | [open.alberta.ca](https://open.alberta.ca/opendata/income-support-aggregated-caseload-data) |

The raw CSV uses a **two-row header** (row 1 is a merged title cell; row 2
contains column names). All data rows report a single
`(month × measure type × measure)` combination in long format.

---

## Data Coverage

The table below shows the temporal extent of each measure type in the dataset.
Understanding this is essential: the `Client Type Level` series (which carries
the BFE/ETW split) only begins in April 2012.

```{r}
#| label: coverage-table

coverage_tbl |>
  rename(
    "Measure Type"  = measure_type,
    "Unique Months" = months,
    "First Month"   = first_month,
    "Last Month"    = last_month
  )
```

> **Key constraint:** The BFE / ETW client-type breakdown is only available from
> **April 2012** onward. The large caseload surge driven by the 2008–2011 Global
> Financial Crisis therefore cannot be attributed to either client type directly.

---

## RQ 1 — What is the Historic Trend of Income Support Caseload?

### Building a continuous 20-year series

Two series together span the full period without gaps:

- **`Total Caseload`** (Apr 2005 – Mar 2012) — single province-wide monthly count.
- **`Client Caseload Total`** within `Client Type Level` (Apr 2012 – Sep 2025) —
  the same concept reported alongside the BFE/ETW split.

Concatenating them gives `r nrow(total_series)` consecutive months.

### Key statistics

```{r}
#| label: stats-callout

cat(
  sprintf("Series start    : %s  (Apr 2005)\n",   scales::comma(total_stats$first_val)),
  sprintf("Series end      : %s  (Sep 2025)\n",   scales::comma(total_stats$last_val)),
  sprintf("All-time minimum: %s  (%s)\n",
          scales::comma(total_stats$min_value), format(total_stats$min_date, "%b %Y")),
  sprintf("All-time maximum: %s  (%s)\n",
          scales::comma(total_stats$max_value), format(total_stats$max_date, "%b %Y")),
  sprintf("Net change      : %+.1f%%  over 20 years\n", total_stats$pct_change)
)
```

::: {.callout-note}
The caseload is **`r scales::comma(total_stats$last_val)`** in September 2025 —
a net change of **`r sprintf("%+.1f%%", total_stats$pct_change)`** from April 2005.
The all-time peak of **`r scales::comma(total_stats$max_value)`**
occurred in **`r format(total_stats$max_date, "%B %Y")`**, driven by the
Great Financial Crisis.
:::

### Chart 1 — Full monthly caseload, 2005–2025

Shaded regions mark the three macro-economic shocks that produced the most
visible movements in the series.

```{r}
#| label: chart-total
#| fig-cap: "Monthly total Income Support caseload, April 2005 – September 2025. Shaded bands show key disruption periods (GFC, oil-price collapse, COVID/CERB)."

chart_total
```

**Narrative:** The series traces *three distinct surge-and-recovery cycles*:

1. **2008–2012 (Global Financial Crisis)** — The sharpest rise in the 20-year
   record. Caseload climbed from roughly 34 000 to a peak near
   **`r scales::comma(total_stats$max_value)`**, then recovered as Alberta's
   labour market rebounded on oil-sands growth.
2. **2015–2017 (Oil price collapse)** — A second, smaller surge as sustained low
   oil prices drove layoffs in the energy sector. Recovery was incomplete before
   COVID arrived.
3. **2020 (COVID / CERB)** — A sharp but *brief* spike driven by pandemic
   unemployment. The federal Canada Emergency Response Benefit partly substituted
   for IS in 2020–21, which may explain the shorter duration compared with the
   GFC surge.

**Post-2021 trend:** Caseload has declined steadily, reaching its lowest level
since the early 2000s by late 2024 and remaining low into 2025.

---

## RQ 2 — How Does the BFE / ETW Breakdown Change the Story?

The `Client Type Level` measure distinguishes two programme streams:

| Stream | Description |
|:-------|:------------|
| **BFE** | Barriers to Full Employment — clients with chronic, long-term barriers (health, caregiving). Higher benefits; lower employment expectation. |
| **ETW – Working** | Expected to Work, currently employed. Receiving income supplementation. |
| **ETW – Available** | Expected to Work, not working but available. Actively seeking employment. |
| **ETW – Unavailable** | Expected to Work, not working and temporarily unavailable (e.g. short-term illness). |

### Chart 2 — BFE vs ETW total caseload

```{r}
#| label: chart-bfe-etw
#| fig-cap: "BFE and total ETW caseload, April 2012 – September 2025."

chart_bfe_etw
```

### Chart 3 — Share of total caseload

```{r}
#| label: chart-share
#| fig-cap: "BFE and ETW as a percentage of combined caseload."

chart_share
```

**Narrative:** The two-line view reveals a *structural divergence* that is
invisible in the aggregate total:

- **BFE** has grown *relative to* ETW over the period. The BFE share rose from
  roughly 45 % in 2012 to well above 50 % by the mid-2020s. This implies the
  IS caseload is increasingly composed of people with more complex, persistent
  barriers — not transient labour-market exits.
- **ETW** fluctuates more with economic conditions, rising sharply during the
  2015–17 oil-price collapse and the 2020 COVID shock, then declining in
  recoveries. This makes ETW a *leading indicator* of labour-market stress.
- During COVID specifically, the ETW spike was *muted* relative to what the
  aggregate total would suggest — consistent with CERB diverting many
  newly-unemployed people away from IS entirely.

### Chart 4 — ETW sub-type detail

```{r}
#| label: chart-etw-detail
#| fig-cap: "ETW caseload broken into three sub-types: Working, Available (for work), and Unavailable."

chart_etw_detail
```

**Narrative:** Within ETW, the **Available** sub-type (not working, looking for
work) drives most of the ETW cyclicality observed in Chart 2. The **Working**
sub-type (income supplementation while employed) has grown slowly but steadily,
reflecting program design changes. The **Unavailable** sub-type is the smallest
and most stable component.

### Chart 5 — Stacked composition view

```{r}
#| label: chart-stacked
#| fig-cap: "Stacked area chart showing the composition of IS caseload by the four client sub-types, April 2012 – September 2025."

chart_stacked
```

---

## Summary Findings

| Question | Finding |
|:---------|:--------|
| **Historic trend** | Three surge-and-recovery cycles over 20 years (GFC, oil-price collapse, COVID). The long-run direction is downward from the GFC peak, with caseload near historic lows in 2024–2025. |
| **BFE vs ETW** | The BFE share of caseload has grown structurally, while ETW is cyclically sensitive. The aggregate trend masks a compositional shift toward more complex cases. |
| **ETW sub-types** | The "Available" sub-type drives ETW cyclicality. The "Working" sub-type has trended upward, suggesting more working-poor households receiving top-up benefits. |

---

## Reproducibility

This document is rendered from a **literate pair**:

- `analysis.R` — all data ingestion, cleaning, and chart construction
- `analysis.qmd` — narrative, tables, and chart display (sources `analysis.R`)

To re-render, run from the `starter-4-sonnet-46/` directory:

```bash
quarto render analysis.qmd
```

```{r}
#| label: session-info
#| code-fold: true
#| code-summary: "Session info"

sessionInfo()
```
