---
title: "Alberta Income Support Caseload: Historic Trends and Client-Type Dynamics"
author: "Literate analysis — Starter 28 (Claude Opus 4.6)"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(scales)
```

## Data Ingestion

The raw CSV from Alberta Open Government uses a two-row header and embeds commas
inside quoted numeric values. We skip the title row, strip the trailing empty
columns, and parse `Value` as a proper numeric.

```{r}
#| label: ingest
raw <- read_csv(
  "../../is-aggregated-data-april-2005-sep-2025.csv",
  skip = 1,                         # drop the title row
  col_types = cols(.default = "c")  # read everything as character first
) |>
  select(1:5) |>                     # keep only meaningful columns
  set_names(c("ref_date", "geography", "measure_type", "measure", "value"))

# Parse the value column: strip spaces and embedded commas, then convert
raw <- raw |>
  mutate(
    value = str_trim(value) |> str_remove_all(",") |> as.numeric()
  )

# Parse the date: format is "YY-Mon" (e.g. "05-Apr", "25-Sep")
raw <- raw |>
  mutate(
    date = parse_date(paste0("01-", ref_date), format = "%d-%y-%b")
  )

glimpse(raw)
```

## Building a Unified Total Caseload Series

The dataset reports total caseload under two different labels depending on the
era. Before April 2012 the measure is called **Total Caseload**; from April 2012
onward it appears as **Client Caseload Total** within the *Client Type Level*
measure type. We stitch these together into one continuous monthly series.

```{r}
#| label: total-series
totals <- raw |>
  filter(
    (measure_type == "Total Caseload" & measure == "Total Caseload") |
    (measure_type == "Client Type Level" & measure == "Client Caseload Total")
  ) |>
  arrange(date)

cat(
  "Total series spans", format(min(totals$date), "%B %Y"),
  "to", format(max(totals$date), "%B %Y"),
  paste0("(", nrow(totals), " months)\n")
)
```

## Research Question 1 — What Is the Historic Trend?

```{r}
#| label: fig-total-trend
#| fig-cap: "Monthly Income Support caseload in Alberta, April 2005 – September 2025"
#| fig-height: 5
ggplot(totals, aes(date, value)) +
geom_line(linewidth = 0.6, colour = "#2c3e50") +
  geom_smooth(method = "loess", span = 0.15, se = FALSE,
              linewidth = 0.9, colour = "#e74c3c", linetype = "dashed") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  labs(
    x = NULL, y = "Cases",
    title = "Alberta Income Support Caseload",
    subtitle = "Monthly total with LOESS smoother (red dashed)"
  ) +
  theme_minimal(base_size = 13)
```

### Narrative

The overall trajectory is far from a simple upward or downward line. Several
distinct phases emerge:

1. **Pre-recession stability (~2005–2008).** Caseloads hovered around
   25,000–27,000, drifting gently with the economic cycle.

2. **2008–09 recession surge.** A sharp climb from roughly 27,000 to over
   38,000 cases — a rise of about 40 % in barely a year — tracks the impact of
   the global financial crisis on Alberta's labour market.

3. **Partial recovery (2010–2014).** Caseloads eased back toward the mid-30,000s
   but never returned to pre-recession levels, suggesting a structural ratchet
   effect.

4. **Oil-price shock (2015–2016).** The collapse in crude prices pushed caseloads
   sharply upward again, peaking around 2017.

5. **COVID-19 (2020–2021).** A fresh spike layered on top of an already elevated
   baseline; the peak in this window pushed past 60,000.

6. **Post-pandemic plateau (~2022–2025).** Caseloads remain in the 58,000–62,000
   range — roughly double the pre-2008 norm — with no clear sign of retreat.

## Research Question 2 — How Does the BFE / ETW Breakdown Change the Story?

The client-type split is available from April 2012 onward. **BFE** (Barriers to
Full Employment) clients face long-term barriers such as chronic health
conditions. **ETW** (Expected to Work) clients are assessed as able to
participate in the labour market — some are already working, some are available
for work, and some are temporarily unavailable.

```{r}
#| label: client-type-data
client_type <- raw |>
  filter(measure_type == "Client Type Level") |>
  mutate(
    group = case_when(
      str_detect(measure, "^BFE")                  ~ "BFE",
      str_detect(measure, "^ETW")                   ~ "ETW",
      str_detect(measure, "Client Caseload Total")  ~ "Total",
      TRUE                                          ~ NA_character_
    )
  ) |>
  filter(!is.na(group))

# Collapse the three ETW sub-categories into a single ETW total per month
bfe_etw <- client_type |>
  filter(group %in% c("BFE", "ETW")) |>
  summarise(value = sum(value), .by = c(date, group))
```

```{r}
#| label: fig-bfe-etw
#| fig-cap: "BFE vs ETW caseload since April 2012"
#| fig-height: 5
ggplot(bfe_etw, aes(date, value, colour = group)) +
  geom_line(linewidth = 0.7) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(values = c("BFE" = "#8e44ad", "ETW" = "#2980b9")) +
  labs(
    x = NULL, y = "Cases", colour = "Client type",
    title = "Income Support by Client Type",
    subtitle = "Barriers to Full Employment (BFE) vs Expected to Work (ETW)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

### Share of caseload

```{r}
#| label: fig-share
#| fig-cap: "BFE share of total caseload over time"
#| fig-height: 4
share <- bfe_etw |>
  pivot_wider(names_from = group, values_from = value) |>
  mutate(bfe_share = BFE / (BFE + ETW))

ggplot(share, aes(date, bfe_share)) +
  geom_line(linewidth = 0.7, colour = "#8e44ad") +
  geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey50") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = NULL, y = "BFE share",
    title = "BFE as a Proportion of Total Caseload"
  ) +
  theme_minimal(base_size = 13)
```

### Stacked area view

```{r}
#| label: fig-stacked
#| fig-cap: "Stacked area: composition of Income Support caseload"
#| fig-height: 5
ggplot(bfe_etw, aes(date, value, fill = group)) +
  geom_area(alpha = 0.8) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("BFE" = "#8e44ad", "ETW" = "#2980b9")) +
  labs(
    x = NULL, y = "Cases", fill = "Client type",
    title = "Income Support Caseload — Stacked by Client Type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

### ETW sub-category detail

```{r}
#| label: fig-etw-detail
#| fig-cap: "ETW sub-categories over time"
#| fig-height: 5
etw_detail <- client_type |>
  filter(group == "ETW") |>
  mutate(
    sub = case_when(
      str_detect(measure, "Working Total$") &
        !str_detect(measure, "Not Working")        ~ "Working",
      str_detect(measure, "Available for Work")     ~ "Not Working\n(Available)",
      str_detect(measure, "Unavailable for Work")   ~ "Not Working\n(Unavailable)",
      TRUE                                          ~ NA_character_
    )
  ) |>
  filter(!is.na(sub))

ggplot(etw_detail, aes(date, value, colour = sub)) +
  geom_line(linewidth = 0.7) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    x = NULL, y = "Cases", colour = "ETW status",
    title = "Expected-to-Work Clients by Employment Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

### Narrative — What the breakdown reveals

The aggregate trend masks two fundamentally different dynamics:

- **BFE growth is steady and structural.** The BFE count has climbed from about
  16,000 in 2012 to roughly 19,000 by 2025 — a gradual, monotonic increase that
  barely flinches during economic swings. This reflects an expanding population
  of Albertans with chronic health or life-circumstance barriers, unrelated to
  business-cycle conditions.

- **ETW is the volatile, cyclically sensitive component.** Almost all of the
  dramatic swings visible in the total series — the oil-price spike, the
  COVID-19 surge — originate in the ETW population. During downturns,
  ETW caseloads balloon as newly unemployed workers enter Income Support; during
  recoveries they partially fall back.

- **BFE's share has been declining** despite its absolute growth, because the
  ETW population has grown even faster. In 2012 BFE represented nearly half the
  caseload; by 2025 it accounts for roughly 31 %.

- **Within ETW**, the *Not Working — Available for Work* sub-group dominates and
  drives the cyclical volatility. The *Working* sub-group remains small,
  suggesting that relatively few Income Support recipients maintain paid
  employment. The *Unavailable for Work* segment has grown notably, forming a
  quasi-structural layer within a category designed for labour-market-ready
  clients.

## Summary

The headline caseload story — a doubling from roughly 27,000 to 61,000 over two
decades — is really two layered stories. A slowly expanding BFE base reflects
structural demand from people facing persistent barriers. Superimposed on that
base, the ETW caseload surges and (partially) recedes with each economic shock,
each time settling at a higher floor than before. Understanding this
compositional difference is essential for policy design: interventions aimed at
reducing BFE caseloads require different levers (health services, disability
supports) than those targeting ETW clients (employment programming, job-market
conditions).
