---
title: "Alberta Income Support: Historic Caseload Trends"
subtitle: "Answering two research questions from 20 years of monthly ALSS data"
author: "Literate analysis — GitHub Copilot (Claude Sonnet 4.6)"
date: "April 2026"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 10
    fig-height: 5
    df-print: paged
    highlight-style: github
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Core tidyverse packages used throughout this analysis
library(readr)      # CSV ingestion
library(dplyr)      # Data manipulation
library(tidyr)      # Reshaping (pivot_wider / pivot_longer)
library(lubridate)  # Date parsing
library(scales)     # Number formatting for axes
library(plotly)     # Interactive HTML charts
```

## Research Questions

This document is a reproducible, literate analysis of the Alberta Income Support
caseload dataset. It directly addresses two questions:

1. **What is the historic trend of Income Support caseload?**
2. **How does breaking it down by BFE and ETW change the story?**

---

## Data Source

| Field | Value |
|:------|:------|
| **Publisher** | Alberta Assisted Living and Social Services (ALSS), Government of Alberta |
| **Dataset** | Income Support Caseload |
| **Coverage** | April 2005 – September 2025 (246 calendar months) |
| **Licence** | [Open Government Licence – Alberta](https://open.alberta.ca/licence) |
| **Portal** | [open.alberta.ca](https://open.alberta.ca/opendata/income-support-aggregated-caseload-data) |

The raw CSV uses a **two-row header** — row 1 is a merged title cell, row 2 is
the actual column header. All data rows report a single
`(month × measure type × measure)` combination (long format).

---

## Ingestion and Parsing

### Reading the raw file

We skip the two-header rows and assign column names manually. Nine trailing
empty columns are present in every row; they are dropped immediately.

```{r}
#| label: ingest

DATA_PATH <- "../../is-aggregated-data-april-2005-sep-2025.csv"

raw <- read_csv(
  DATA_PATH,
  col_names = c(
    "ref_date", "geography", "measure_type", "measure", "value",
    paste0("drop_", 1:9)   # nine trailing empty columns
  ),
  skip           = 2,      # skip title row + header row
  show_col_types = FALSE
)

# Drop the empty trailing columns and any rows where value is blank
raw <- raw |>
  select(ref_date, geography, measure_type, measure, value) |>
  filter(!is.na(value), trimws(value) != "")
```

### Parsing dates

Dates are encoded as `YY-Mon` (e.g., `05-Apr`, `20-Jan`). Two-digit years with
value ≤ 25 are assumed to be 2000s; all others would be 1900s. We parse to a
first-of-month `Date` for clean time-series handling.

```{r}
#| label: parse-dates

parse_ref_date <- function(x) {
  x     <- trimws(x)
  parts <- strsplit(x, "-")
  yr    <- as.integer(vapply(parts, `[[`, character(1), 1))
  mon   <- vapply(parts, `[[`, character(1), 2)

  full_year <- ifelse(yr <= 25L, 2000L + yr, 1900L + yr)

  as.Date(
    paste(full_year, mon, "01"),
    format = "%Y %b %d"
  )
}

raw <- raw |>
  mutate(
    date  = parse_ref_date(ref_date),
    value = as.numeric(gsub("[^0-9.]", "", value))
  )

cat(
  sprintf("Rows after cleaning: %d\n", nrow(raw)),
  sprintf("Date range: %s → %s\n",
          format(min(raw$date), "%b %Y"),
          format(max(raw$date), "%b %Y")),
  sprintf("Distinct measure types: %s\n",
          paste(unique(raw$measure_type), collapse = ", "))
)
```

### Available measure types and their coverage

Understanding the temporal extent of each measure type is essential: the
`Total Caseload` series covers 2005–2012, while the richer `Client Type Level`
breakdown (BFE/ETW) only begins in April 2012.

```{r}
#| label: coverage-table

raw |>
  group_by(measure_type) |>
  summarise(
    months       = n_distinct(date),
    first_month  = format(min(date), "%b %Y"),
    last_month   = format(max(date), "%b %Y"),
    .groups = "drop"
  ) |>
  arrange(first_month) |>
  rename(
    "Measure Type"  = measure_type,
    "Unique Months" = months,
    "First Month"   = first_month,
    "Last Month"    = last_month
  )
```

> **Key finding:** The BFE/ETW client-type breakdown is only available from
> **April 2012** onward. The 2008–2011 recession surge therefore cannot be
> attributed directly to either client type.

---

## RQ 1 — What is the Historic Trend of Income Support Caseload?

### Building the full series (2005–2025)

Two overlapping series together span the full period:

- **`Total Caseload`** (Apr 2005 – Mar 2012): a single province-wide monthly figure.
- **`Client Caseload Total`** within `Client Type Level` (Apr 2012 – Sep 2025):
  the same concept, reported alongside the BFE/ETW breakdown.

We concatenate them into one continuous series. Where both exist for the same
month (there is no overlap here — the `Total Caseload` series ends exactly where
`Client Type Level` begins), the latter takes precedence.

```{r}
#| label: build-total-series

total_series <- bind_rows(
  raw |>
    filter(measure_type == "Total Caseload") |>
    select(date, value),

  raw |>
    filter(
      measure_type == "Client Type Level",
      measure      == "Client Caseload Total"
    ) |>
    select(date, value)
) |>
  distinct(date, .keep_all = TRUE) |>
  arrange(date)

cat(sprintf(
  "Combined series: %d months (%s → %s)\n",
  nrow(total_series),
  format(min(total_series$date), "%b %Y"),
  format(max(total_series$date), "%b %Y")
))
```

### Key statistics

```{r}
#| label: stats-total

stats <- total_series |>
  summarise(
    min_value  = min(value),
    min_date   = date[which.min(value)],
    max_value  = max(value),
    max_date   = date[which.max(value)],
    first_val  = first(value),
    last_val   = last(value),
    pct_change = round((last(value) - first(value)) / first(value) * 100, 1)
  )

cat(
  sprintf("All-time minimum : %s  (%s)\n",
          comma(stats$min_value), format(stats$min_date, "%b %Y")),
  sprintf("All-time maximum : %s  (%s)\n",
          comma(stats$max_value), format(stats$max_date, "%b %Y")),
  sprintf("Series start     : %s  (Apr 2005)\n", comma(stats$first_val)),
  sprintf("Series end       : %s  (Sep 2025)\n", comma(stats$last_val)),
  sprintf("Net change       : %+.1f%%  over 20 years\n", stats$pct_change)
)
```

### Chart 1 — Full monthly caseload, 2005–2025

Shaded regions mark the three macro-economic disruptions that produced the
largest movements in the series.

```{r}
#| label: chart-total
#| fig-cap: "Monthly total Income Support caseload, April 2005 – September 2025. Shaded bands show key economic disruption periods."

# Annotation bands: start, end, label, fill colour
bands <- tibble(
  xstart = as.Date(c("2008-10-01", "2015-01-01", "2020-04-01")),
  xend   = as.Date(c("2011-12-01", "2016-12-01", "2021-08-01")),
  label  = c("GFC", "Oil price\ncollapse", "COVID /\nCERB"),
  fill   = c("rgba(255,152,0,0.15)", "rgba(255,152,0,0.15)", "rgba(33,150,243,0.12)")
)

p1 <- plot_ly() |>
  # Shaded annotation bands
  add_ribbons(
    data       = total_series,
    x          = ~date,
    ymin       = 0,
    ymax       = max(total_series$value) * 1.05,
    color      = I("rgba(255,152,0,0.12)"),
    showlegend = FALSE,
    hoverinfo  = "none",
    name       = "GFC"
  )

# Manually add each band as a shape (plotly layout shapes)
shapes <- lapply(seq_len(nrow(bands)), function(i) {
  list(
    type      = "rect",
    xref      = "x", yref   = "paper",
    x0        = bands$xstart[i], x1 = bands$xend[i],
    y0        = 0, y1 = 1,
    fillcolor = bands$fill[i],
    line      = list(width = 0),
    layer     = "below"
  )
})

annotations <- lapply(seq_len(nrow(bands)), function(i) {
  list(
    x          = bands$xstart[i] + as.numeric(bands$xend[i] - bands$xstart[i]) / 2,
    y          = 0.97,
    xref       = "x", yref  = "paper",
    text       = bands$label[i],
    showarrow  = FALSE,
    font       = list(size = 10, color = "rgba(80,80,80,0.8)"),
    align      = "center"
  )
})

p1 <- plot_ly(total_series, x = ~date, y = ~value, type = "scatter",
              mode = "lines",
              line = list(color = "#1a6eb5", width = 2),
              fill = "tozeroy", fillcolor = "rgba(26,110,181,0.08)",
              name = "Total Caseload",
              hovertemplate = "<b>%{x|%b %Y}</b><br>Caseload: %{y:,.0f}<extra></extra>") |>
  layout(
    title       = list(text = "Alberta Income Support — Total Caseload", font = list(size = 15)),
    xaxis       = list(title = "", showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)"),
    yaxis       = list(title = "Caseload (clients)", tickformat = ",.0f",
                      showgrid = TRUE, gridcolor = "rgba(0,0,0,0.06)"),
    shapes      = shapes,
    annotations = annotations,
    legend      = list(orientation = "h"),
    hovermode   = "x unified",
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  )

p1
```

### Narrative: five phases of the trend

| Phase | Period | Direction | Driver |
|:------|:-------|:----------|:-------|
| 1 — Boom | Apr 2005 – Sep 2008 | ↓ Falling | Alberta oil boom: tight labour market drew potential clients into employment |
| 2 — GFC surge | Oct 2008 – Jun 2010 | ↑ +48% | Global Financial Crisis; local recession; rapid case intake |
| 3 — Partial recovery | 2011 – 2014 | → Plateau | Recovery stalled; caseload stabilised in low-to-mid 30,000s |
| 4 — Oil crash & climb | Jan 2015 – Jan 2020 | ↑ +76% | Oil price collapse; caseload nearly doubled to all-time record of **62,021** |
| 5 — COVID drop | Apr 2020 – Aug 2021 | ↓ −34% | Federal CERB/CRB replaced provincial IS for many able-bodied clients |
| 6 — Post-pandemic surge | Sep 2021 – Sep 2025 | ↑ +49% | Federal supports wound down; caseload rebounding rapidly toward prior record |

As of September 2025, the caseload stands at **61,061** — 98% of the January 2020 all-time record and rising.

---

## RQ 2 — How Does the BFE / ETW Breakdown Change the Story?

### Client type definitions

- **BFE (Barriers to Full Employment):** Clients with significant, often long-term barriers — chronic health conditions (>6 months), complex caregiving situations, or other circumstances that preclude full-time employment. Generally receive higher benefit levels.
- **ETW (Expected to Work):** Clients assessed as able to participate in employment. Subdivided into three subcategories: currently *Working* (low-earnings supplement), *Available for work but not working*, and *Temporarily unavailable*.

### Building the client-type series

```{r}
#| label: build-split-series

client_type_measures <- c(
  "BFE - Total",
  "ETW - Working Total",
  "ETW - Not Working (Available for Work) Total",
  "ETW - Not Working (Unavailable for Work) Total",
  "Client Caseload Total"
)

split_series <- raw |>
  filter(
    measure_type == "Client Type Level",
    measure      %in% client_type_measures
  ) |>
  select(date, measure, value) |>
  # Tidy the measure names for use as column names / labels
  mutate(
    measure_label = case_match(
      measure,
      "BFE - Total"                                       ~ "bfe",
      "ETW - Working Total"                               ~ "etw_working",
      "ETW - Not Working (Available for Work) Total"      ~ "etw_avail",
      "ETW - Not Working (Unavailable for Work) Total"    ~ "etw_unavail",
      "Client Caseload Total"                             ~ "total"
    )
  ) |>
  select(-measure) |>
  pivot_wider(names_from = measure_label, values_from = value) |>
  arrange(date) |>
  mutate(
    etw_total = etw_working + etw_avail + etw_unavail,
    bfe_share = round(bfe / total * 100, 1),
    etw_share = round(etw_total / total * 100, 1)
  )

cat(sprintf(
  "Split series: %d months (%s → %s)\n",
  nrow(split_series),
  format(min(split_series$date), "%b %Y"),
  format(max(split_series$date), "%b %Y")
))
```

### Annual summary: BFE vs. ETW

```{r}
#| label: annual-split-table

split_series |>
  mutate(year = year(date)) |>
  group_by(year) |>
  summarise(
    across(c(total, bfe, etw_total, etw_avail, etw_unavail, etw_working, bfe_share),
           ~ round(mean(.x))),
    .groups = "drop"
  ) |>
  rename(
    "Year"              = year,
    "Total"             = total,
    "BFE"               = bfe,
    "ETW Total"         = etw_total,
    "ETW Avail."        = etw_avail,
    "ETW Unavail."      = etw_unavail,
    "ETW Working"       = etw_working,
    "BFE Share (%)"     = bfe_share
  )
```

### Chart 2 — BFE vs. total ETW over time

The plot below immediately reveals the fundamental divergence: BFE is stable;
ETW is volatile and growing.

```{r}
#| label: chart-bfe-etw
#| fig-cap: "Monthly BFE and ETW caseload, April 2012 – September 2025. BFE is stable; ETW drives every major swing."

shapes_split <- lapply(list(
  list(x0 = "2015-01-01", x1 = "2016-12-01", fill = "rgba(255,152,0,0.12)", lbl = "Oil crash"),
  list(x0 = "2020-04-01", x1 = "2021-08-01", fill = "rgba(33,150,243,0.10)", lbl = "COVID/CERB")
), function(b) {
  list(type="rect", xref="x", yref="paper",
       x0=as.Date(b$x0), x1=as.Date(b$x1),
       y0=0, y1=1, fillcolor=b$fill, line=list(width=0), layer="below")
})

ann_split <- lapply(list(
  list(x0="2015-01-01", x1="2016-12-01", lbl="Oil crash"),
  list(x0="2020-04-01", x1="2021-08-01", lbl="COVID/CERB")
), function(b) {
  xs <- as.Date(b$x0); xe <- as.Date(b$x1)
  list(x=xs + as.numeric(xe-xs)/2, y=0.97, xref="x", yref="paper",
       text=b$lbl, showarrow=FALSE,
       font=list(size=10, color="rgba(80,80,80,0.8)"))
})

plot_ly(split_series) |>
  add_trace(x = ~date, y = ~bfe,
            type = "scatter", mode = "lines",
            name = "BFE – Total",
            line = list(color = "#c0392b", width = 2.5),
            fill = "tozeroy", fillcolor = "rgba(192,57,43,0.08)",
            hovertemplate = "<b>%{x|%b %Y}</b><br>BFE: %{y:,.0f}<extra></extra>") |>
  add_trace(x = ~date, y = ~etw_total,
            type = "scatter", mode = "lines",
            name = "ETW – Total",
            line = list(color = "#2980b9", width = 2.5),
            fill = "tozeroy", fillcolor = "rgba(41,128,185,0.08)",
            hovertemplate = "<b>%{x|%b %Y}</b><br>ETW: %{y:,.0f}<extra></extra>") |>
  layout(
    title       = list(text = "BFE vs. ETW — Monthly Caseload (Apr 2012 – Sep 2025)",
                       font = list(size = 15)),
    xaxis       = list(title = "", gridcolor = "rgba(0,0,0,0.06)"),
    yaxis       = list(title = "Clients", tickformat = ",.0f",
                       gridcolor = "rgba(0,0,0,0.06)"),
    shapes      = shapes_split,
    annotations = ann_split,
    legend      = list(orientation = "h", y = -0.12),
    hovermode   = "x unified",
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  )
```

### Chart 3 — BFE share of total caseload (%)

A declining share confirms the caseload is becoming increasingly ETW-dominant
over the observation period.

```{r}
#| label: chart-bfe-share
#| fig-height: 3.5
#| fig-cap: "BFE share of total caseload, monthly. Falling from ~50% in 2012 to ~31% in September 2025."

plot_ly(split_series, x = ~date, y = ~bfe_share,
        type = "scatter", mode = "lines",
        line = list(color = "#c0392b", width = 2),
        fill = "tozeroy", fillcolor = "rgba(192,57,43,0.08)",
        name = "BFE share (%)",
        hovertemplate = "<b>%{x|%b %Y}</b><br>BFE share: %{y:.1f}%<extra></extra>") |>
  layout(
    title  = list(text = "BFE Share of Total Caseload (%)", font = list(size = 15)),
    xaxis  = list(title = "", gridcolor = "rgba(0,0,0,0.06)"),
    yaxis  = list(title = "BFE share (%)", range = c(25, 60),
                  ticksuffix = "%", gridcolor = "rgba(0,0,0,0.06)"),
    shapes = c(
      shapes_split,
      list(list(type="line", xref="paper", yref="y",
                x0=0, x1=1, y0=50, y1=50,
                line=list(color="rgba(108,117,125,0.5)", width=1,
                          dash="dash")))
    ),
    hovermode     = "x unified",
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  )
```

### Chart 4 — ETW subcategory breakdown

Within ETW, three subcategories tell quite different stories. This chart lets
you compare them in stacked or individual form.

```{r}
#| label: chart-etw-sub
#| fig-cap: "ETW subcategories, monthly, April 2012 – September 2025. 'Available for work' is overwhelmingly the largest and fastest-growing group."

plot_ly(split_series) |>
  add_trace(x = ~date, y = ~etw_avail,
            type = "scatter", mode = "lines", stackgroup = "etw",
            name = "Not Working – Available",
            line = list(color = "#2980b9"),
            fillcolor = "rgba(41,128,185,0.5)",
            hovertemplate = "<b>%{x|%b %Y}</b><br>Available: %{y:,.0f}<extra></extra>") |>
  add_trace(x = ~date, y = ~etw_unavail,
            type = "scatter", mode = "lines", stackgroup = "etw",
            name = "Not Working – Unavailable",
            line = list(color = "#8e44ad"),
            fillcolor = "rgba(142,68,173,0.5)",
            hovertemplate = "<b>%{x|%b %Y}</b><br>Unavailable: %{y:,.0f}<extra></extra>") |>
  add_trace(x = ~date, y = ~etw_working,
            type = "scatter", mode = "lines", stackgroup = "etw",
            name = "Working (earnings supplement)",
            line = list(color = "#27ae60"),
            fillcolor = "rgba(39,174,96,0.5)",
            hovertemplate = "<b>%{x|%b %Y}</b><br>Working: %{y:,.0f}<extra></extra>") |>
  layout(
    title       = list(text = "ETW Subcategories — Stacked Monthly Caseload",
                       font = list(size = 15)),
    xaxis       = list(title = "", gridcolor = "rgba(0,0,0,0.06)"),
    yaxis       = list(title = "Clients", tickformat = ",.0f",
                       gridcolor = "rgba(0,0,0,0.06)"),
    shapes      = shapes_split,
    annotations = ann_split,
    legend      = list(orientation = "h", y = -0.15),
    hovermode   = "x unified",
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  )
```

---

## Synthesis: Answering the Research Questions

### RQ 1 — What is the historic trend?

The 20-year record is one of sustained, structural growth punctuated by cyclical
shocks. Starting from a boom-era low of **24,205** (October 2006), total caseload
has grown by over 150% to **61,061** (September 2025). The net trajectory is
unmistakably upward — even after recovering from the COVID dip, the caseload is
now approaching the all-time record of 62,021 set in January 2020.

Three macro-shocks produced large step-changes: the **Global Financial Crisis**
(+48%, 2008–2010), the **oil price collapse** (+76%, 2015–2020), and the
**post-pandemic rebound** (under way since 2021). Notably, the COVID period
produced a *downward* shock — an anomaly explained by federal emergency income 
supports (CERB/CRB) temporarily displacing provincial Income Support.

### RQ 2 — How does BFE/ETW change the story?

The single trend line conceals a fundamental division:

```{r}
#| label: final-stats
#| echo: false

first_row <- split_series |> filter(date == min(date))
last_row  <- split_series |> filter(date == max(date))
bfe_peak  <- split_series |> filter(bfe == max(bfe))
etw_peak  <- split_series |> filter(etw_total == max(etw_total))

tibble(
  Segment      = c("BFE", "ETW Total", "ETW – Available for Work"),
  `Apr 2012`   = c(
    comma(first_row$bfe),
    comma(first_row$etw_total),
    comma(first_row$etw_avail)
  ),
  `Sep 2025`   = c(
    comma(last_row$bfe),
    comma(last_row$etw_total),
    comma(last_row$etw_avail)
  ),
  `Peak`       = c(
    paste0(comma(bfe_peak$bfe), " (", format(bfe_peak$date, "%b %Y"), ")"),
    paste0(comma(etw_peak$etw_total), " (", format(etw_peak$date, "%b %Y"), ")"),
    "—"
  ),
  `Growth (2012–2025)` = c(
    paste0("+", round((last_row$bfe - first_row$bfe) / first_row$bfe * 100), "%"),
    paste0("+", round((last_row$etw_total - first_row$etw_total) / first_row$etw_total * 100), "%"),
    paste0("+", round((last_row$etw_avail - first_row$etw_avail) / first_row$etw_avail * 100), "%")
  )
) 
```

**BFE** represents relatively stable, structural need. Its modest growth
(~+18% over 13 years) likely reflects demographic trends and policy thresholds
rather than economic cycles. It is now a *shrinking share* of the caseload —
down from ~50% in 2012 to ~31% in September 2025.

**ETW** is the sole driver of large-scale caseload variation. Every recession,
every oil-price downturn, and every pandemic-recovery curve is an ETW story.
Within ETW, the *Available for Work* subcategory — clients ready to work but
unable to find it — has grown nearly four-fold (~7,200 to ~27,000), making it
the single largest contributor to the 20-year upward trend.

**Policy implication:** A caseload that is now 69% ETW is fundamentally a
**labour market accessibility problem**, not primarily a chronic-need or
disability story. Caseload responds sharply to economic conditions (booms pull
it down; recessions push it up), and the post-pandemic trajectory suggests that
recovery of labour market inclusion has not kept pace with the withdrawal of
emergency federal income supports.

---

## Appendix: Session Information

```{r}
#| label: session-info
#| code-fold: false

cat("R version:", R.version$version.string, "\n")
cat("Platform :", R.version$platform, "\n\n")
pkgs <- c("readr","dplyr","tidyr","lubridate","scales","plotly")
ip   <- as.data.frame(installed.packages()[pkgs, c("Package","Version")])
rownames(ip) <- NULL
print(ip)
```
